The lattice of acyclic pipe dreams

### 2023-01-11 acyclicFacets/draft/latticePipeDreams.tex
section 3 completed

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}

@@

% pipe dreams
\input{pipeDreamsTikz}
% cross
%\newcommand{\cross}[1][black]{\raisebox{.1cm}{\pipeDreamMonoColor{#1}{black}{c}}}
\newcommand{\cross}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{cross}}}
\newcommand{\crossBiColor}[2]{\raisebox{.1cm}{\pipeDreamBiColor{#1}{#2}{black}{c/l/r}}}
\newcommand{\NScross}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{NScross}}}
\newcommand{\WEcross}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{WEcross}}}
% elbows
%\newcommand{\elbow}[1][black]{\raisebox{.1cm}{\pipeDreamMonoColor{#1}{black}{e}}}
\newcommand{\elbow}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{elbow}}}
\newcommand{\elbowBiColor}[2]{\raisebox{.1cm}{\pipeDreamBiColor{#1}{#2}{black}{e/l/r}}}
%\newcommand{\SEelbow}[1][black]{\raisebox{.1cm}{\pipeDreamBiColor{white}{#1}{black}{e/l/r}}}
\newcommand{\SEelbow}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{SEelbow}}}
%\newcommand{\WNelbow}[1][black]{\raisebox{.1cm}{\pipeDreamBiColor{#1}{white}{black}{e/l/r}}}
\newcommand{\WNelbow}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{WNelbow}}}
\newcommand{\waving}{\mathsf{wav}}%{\raisebox{-.2cm}{\begin{tikzpicture}[scale=1]\pipeDreamBiColor{.2}{(0,0)}{black}{white}{{t/r/l},{t/l/l},{t/l/r}}\end{tikzpicture}}}
\newcommand{\transpose}{\mathsf{transp}}
\newcommand{\pipeDreams}{\Pi} % set of pipe dreams
\newcommand{\HP}{{\bk\pipeDreams}} % Hopf Algebra of pipe dreams
\newcommand{\reversingPipeDreams}{\Omega} % reversing pipe dreams
\newcommand{\contact}{^\#} % contact graph
\newcommand{\se}{\textsc{se}} % southeast
\newcommand{\nw}{\textsc{nw}} % northwest
\newcommand{\insertion}[2]{\mathsf{ins}(#1,#2)} % insertion of permutation #1 in pipe dreams with permutation #2
\newcommand{\duality}{^\star} % duality pipe dreams - triangulations
\newcommand{\acyclicPipeDreams}{\Sigma} % acyclic pipe dreams
\newcommand{\linearExtensions}{\mathcal{L}} % linear extensions
\newcommand{\noninversions}[2]{\mathsf{ninv}(#1,#2)} % non-inversions
\newcommand{\acyclicOrientations}{\Omega} % set of pipe dreams
\newcommand{\recoils}[2]{\mathsf{rec}(#1,#2)} % recoil scheme of a pipe dream
\newcommand{\canopy}[1]{\mathsf{can}(#1)} % canopy of a pipe dream
\newcommand{\greedyPipeDream}{P^{\overleftarrow{gr}}} % greedy facet
\newcommand{\antiGreedyPipeDream}{P^{\overrightarrow{gr}}} % antigreedy facet

@@

@@

We call \defn{pipe dream congruence} the equivalence relation~$\equiv_\omega$ on the weak order interval~$[e,\omega]$, whose equivalence classes are the sets of linear extensions of the pipe dreams of~$\acyclicPipeDreams(\omega)$.
This indeed defines an equivalence relation by \cref{thm:partitionPipeDreams}.
In this section, we prove the following statement.

@@

\begin{example}
\label{exm:Tamari3}
Following \cref{exm:Tamari1,exm:Tamari2}, observe that the congruence~$\equiv_{\rho_n}$ on the permutations of~$\{0, 1, \dots, n, n+1\}$ below~$\rho_n$ corresponds to the sylvester congruence on~$\fS_n$.
The classes of this congruence are the sets of linear extensions of the binary trees (considered as posets, labeled in inorder, and oriented towards their roots).
It can also be defined by the classical rewriting rule~$UikVjW \equiv UkiVjW$ for~$i < j < k$.
\end{example}

@@

In this section, we discuss four further topics on the pipe dream congruence.
In \cref{subsec:sweepingAlgorithm,subsec:insertionAlgorithm}, we describe two algorithms to construct the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ of which a given permutation~$\pi$ is a linear extension.
In \cref{subsec:rewritingRule}, we describe the pipe dream congruence~$\equiv_\omega$ as the transitive closure of a rewriting rule on permutations of~$[e, \omega]$.
Finally, in \cref{subsec:canopy}, we describe the natural coarsening of the pipe dream congruence~$\equiv_\omega$ by the recoil congruence~$\simeq_\omega$.

@@

Our first algorithm to construct the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ with~$\pi \in \linearExtensions(P)$ is a \defn{sweeping algorithm}, inspired by the algorithm to compute greedy pipe dreams~\cite{PilaudPocchiola, PilaudStump-ELlabelings}.
Extensions of this algorithm will be discussed in \cref{..}. \vincent{todo. Was this algorithm explicitly somewhere before? We should cite Jahn-Stump later in the paper.}

@@

Our second algorithm to construct the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ with~$\pi \in \linearExtensions(P)$ is an \defn{insertion algorithm} inspired from~\cite{Pilaud-BrickAlgebra} and similar to the insertion in binary search trees.

@@

\begin{lemma}
\label{lem:rectangleInsertionAlgorithm}
For any~$1 \le t \le r \le n$, the free southeast elbows in the rectangle~$[\pi(r)] \times [\omega^{-1}(\pi(r))]$ just before step~$t$ of the insertion algorithm form a staircase of length
\[
{\# \bigset{s < t}{\pi(s) < \pi(r) \text{ and } \omega^{-1}(\pi(s)) < \omega^{-1}(\pi(r))}}.
\]
\end{lemma}

\begin{proof}
Denote by~$R$ the rectangle~$[\pi(r)] \times [\omega^{-1}(\pi(r))]$.
The proof works by induction on~$t$.
Before step~$1$, there is no southeast elbow in~$R$.
Assume now that just before step~$t$, the free southeast elbows in~$R$ form a staircase~$e_1, \dots, e_k$ with~$k = \# \bigset{{s < t}}{{\pi(s) < \pi(r)} \text{ and } {\omega^{-1}(\pi(s)) < \omega^{-1}(\pi(r))}}$.
At step~$t$, we insert a new pipe~$\pi(t)$ which enters at~$\pi(t)$ and exits at~$\omega^{-1}(\pi(t))$.
Let~${0 \le i < j \le k+1}$ be such that~$e_{i+1}, \dots, e_{j-1}$ are the northwest elbows of pipe~$\pi(t)$ in~$R$, and let~$e'_1, \dots, e'_\ell$ be the southeast elbows of pipe~$\pi(t)$ in~$R$.
Hence, the free southeast elbows in~$R$ after step~$t$ form the sequence~$e_1, \dots, e_i, e'_1, \dots, e'_\ell, e_j, \dots, e_k$.
We thus just need to see that
\begin{itemize}
\item $e_i$ is southwest of~$e'_1$ and~$e'_\ell$ is southwest of~$e_j$,
\item $\ell = j-i$ if~$\pi(s) < \pi(t)$ and~$\omega^{-1}(\pi(s)) > \omega^{-1}(\pi(t))$, and $\ell = j-i-1$ otherwise.
\end{itemize}
This immediately follows considering the four possible cases:
\begin{itemize}
\item If~$\pi(t) < \pi(r)$ and~$\omega^{-1}(\pi(t)) < \omega^{-1}(\pi(r))$, then the pipe~$\pi(t)$ is completely contained in~$R$, hence it has one more northwest elbow than southeast elbow in~$R$.
\item If~$\pi(s) < \pi(t)$ and~$\omega^{-1}(\pi(s)) > \omega^{-1}(\pi(t))$, then the pipe~$\pi(s)$ enters and exits~$R$ with an horizontal step, hence it has as many northwest and southeast elbows in~$R$.
\item If~$\pi(s) > \pi(t)$ and~$\omega^{-1}(\pi(s)) < \omega^{-1}(\pi(t))$, then the pipe~$\pi(s)$ enters and exits~$R$ with a vertical step, hence it has as many northwest and southeast elbows in~$R$.
\item If~$\pi(s) > \pi(t)$ and~$\omega^{-1}(\pi(s)) > \omega^{-1}(\pi(t))$, then setting~$i \eqdef \pi(t)$ and~$j \eqdef \pi(s)$, we have~${i < j}$ and~$\pi^{-1}(i) > \pi^{-1}(j)$ while~$\omega^{-1}(i) < \omega^{-1}(j)$, which contradicts our assumption that~$\pi < \omega$ in weak order. \qedhere
\end{itemize}
\end{proof}

@@

\begin{proof}
%Fix~$t \in [n]$.
Observe first that if there is~$s > t$ such that~$\pi(s) < \pi(t)$ and~$\omega^{-1}(\pi(s)) < \omega^{-1}(\pi(t))$, then setting~$i \eqdef \pi(s)$ and~$j \eqdef \pi(t)$, we have~$i < j$ and~$\pi^{-1}(i) > \pi^{-1}(j)$ while~$\omega^{-1}(i) < \omega^{-1}(j)$, which contradicts our assumption that~$\pi < \omega$ in weak order.
Hence, we obtain that
\[
\noninversions{\omega}{\pi(t)} = \# \bigset{s < t}{\pi(s) < \pi(t) \text{ and } \omega^{-1}(\pi(s)) < \omega^{-1}(\pi(t))}.
\]
Hence, the result follows from~\cref{lem:rectangleInsertionAlgorithm} applied to~$t = r$.
\end{proof}

@@

%\begin{corollary}
%For any~$\pi \le \omega$, the insertion algorithm creates a pipe dream of~$\pipeDreams(\omega)$.
%\end{corollary}

@@

%\begin{proposition}
%\label{prop:fibersInsertion}
%For any pipe dream~$P \in \acyclicPipeDreams(\omega)$,  the permutations~$\pi \le \omega$ such that~$\insertion{\pi}{\omega} = P$ are precisely the linear extensions of the contact graph~$P\contact$ of~$P$.
%In particular, $\insertion{\cdot}{\omega}$ is a surjection from~$[e, \omega]$ to~$\acyclicPipeDreams(\omega)$.
%\end{proposition}
%
%\begin{proof}
%By construction of the insertion algorithm, all southeast contacts of the pipe~$\pi(t)$ inserted at step~$t$ are in contact with northwest contacts of pipes~$\pi(s)$ inserted at steps~$s < t$.
%In other words, all edges in the contact graph of~$\insertion{\pi}{\omega}$ are of the form~$\pi(s) \to \pi(t)$ for some~$s < t$.
%It follows that the permutation~$\pi$ is a linear extension of~$\insertion{\pi}{\omega}$.
%Reciprocally, consider a linear extension~$\pi$ of a pipe dream~$P$.
%Then~$\insertion{\pi}{\omega}$ and~$P$ have a common linear extension~$\pi$, so that~$\insertion{\pi}{\omega} = P$ by \cref{prop:partition}.
%\end{proof}
%
%In other words, the sets~$\set{\linearExtensions(P)}{P \in \acyclicPipeDreams(\omega)}$ are the fibers of the surjective insertion map~$\insertion{\cdot}{\omega}$.
%Note that this concludes the proof of the surjectivity left open in \cref{prop:partition}.

@@

\vincent{I am still working below this line.}

Recall from \cref{exm:Tamari3} that the sylvester congruence can be defined as the transitive closure of the classical rewriting rule~$UikVjW \equiv UkiVjW$ for~$i < j < k$ (as usual, we write the permutations of~$\fS_n$ as words in one-line notation).
We now describe a similar rewriting rule for the pipe dream congruence~$\equiv_\omega$.

\begin{proposition}
\label{prop:rewritingRule}
On the interval~$[e,\omega]$ of the weak order, the pipe dream congruence~$\equiv_\omega$ coincides with the transitive closure of the rewriting rule~$U ik V \equiv_\omega U ki V$ where~$i < k$ are elements of~$[n]$ while~$U, V$ are (possibly empty) words on~$[n]$ such that
\[
\#\set{j \in U}{i < j < k \text{ and } \omega^{-1}(i) > \omega^{-1}(j) > \omega^{-1}(k)} \ge \#\set{j \in U}{j < i \text{ and } \omega^{-1}(j) < \omega^{-1}(k)}.
%\#\set{j \in U}{i < j < k \text{ and } \omega^{-1}(i) > \omega^{-1}(j) > \omega^{-1}(k)} \ge \#\set{j < i}{\omega^{-1}(j) < \omega^{-1}(k)},
\]
\vincent{Note that $b \in U$ on the right hand side by \cref{lem:interval}.}
% The letters~$j \in U$ such that~$i < j < k$ while~$\omega^{-1}(i) > \omega^{-1}(j) > \omega^{-1}(k)$ are called \defn{pipe dream congruence witnesses} for the exchange of~$i$ and~$k$.
%\vincent{Another option to write this is $d \ge u$ where $d \eqdef \set{b \in U}{b > a}$ and~$u \eqdef \set{b \in U}{\omega^{-1}(b) < \omega^{-1}(c)}$.
%Or $dr \ge \ell u$ where $dr \eqdef \set{b \in U}{b > a \text{ and } \omega^{-1}(b) > \omega^{-1}(c)}$ and~$\ell u \eqdef \set{b \in U}{b < a \text{ and } \omega^{-1}(b) < \omega^{-1}(c)}$.}
\end{proposition}

\begin{proposition}
\label{prop:congruence}
For any~$\pi, \pi' \in [e,\omega]$, we have~${\pi \equiv_\omega \pi' \!\iff\! \insertion{\pi}{\omega} = \insertion{\pi'}{\omega} \!\iff\! \pi}$~and~$\pi'$ are linear extensions of a same pipe dream.
\end{proposition}

\begin{proof}
By \cref{prop:fibersInsertion}, each fiber of~$\insertion{\cdot}{\omega}$ gathers the linear extensions of a pipe dream of~$\pipeDreams(\omega)$.
Since the set of linear extensions of a poset is connected by simple transpositions, we just need to show that~$\pi \equiv_\omega \pi' \!\iff\! \insertion{\pi}{\omega} = \insertion{\pi'}{\omega}$ for any two permutations~$\pi = UacV$ and~$\pi' = UcaV$ of~$[e,\omega]$ which differ by the inversion of two consecutive values.

Let~$t = \pi^{-1}(a) = \pi'{-1}(c)$.
At step~$t$ of the algorithm presented in the previous section, we have inserted the word~$U$.
At that moment, let~$X_t$ (resp.~$X'_t$, resp~$Y_t$) denote the set of free northwest contacts in the rectangle~$[a] \times [\omega^{-1}(a)]$ (resp.~$[c] \times [\omega^{-1}(c)]$, resp.~$[a] \times [\omega^{-1}(c)]$).
Applying the same argument as in \cref{lem:rectangleInsertionAlgorithm}, we have
\begin{gather*}
\# Y_t = \max\big( \#\set{b \in U}{b < a \text{ and } \omega^{-1}(b) < \omega^{-1}(c)} \hspace{5cm} \\ \hspace{5cm} - \; \#\set{b \in U}{a < b < c \text{ and } \omega^{-1}(a) > \omega^{-1}(b) > \omega^{-1}(c)}, 0\big).
\end{gather*}
It follows that~$U ac V \equiv_\omega U ca V \iff \# Y_t = 0$.
However, $Y_t = 0$ if and only if all contacts of~$X_t$ are in~$[a] \times [\omega^{-1}(c), \omega^{-1}(a)]$ and all contacts of~$X'_t$ are in~$[a,c] \times [\omega^{-1}(c)]$.
The latter is equivalent to the fact that the insertion of~$a$ and~$c$ commute.
We conclude that
\[
U ac V \equiv_\omega U ca V \iff \insertion{UacV}{\omega} = \insertion{UcaV}{\omega},
\]
which concludes the proof since the set of linear extensions of a poset is connected by simple transpositions.
\end{proof}

\begin{proposition}
\label{prop:patternAvoiding}
The minimal (resp.~maximal) permutations in the partition
\(
{[e,\omega] = \bigsqcup_{P \in \acyclicPipeDreams(\omega)} \linearExtensions(P)}
\)
are the permutations avoiding the patterns~$b_1 - \cdots - b_k - ca$ (resp.~$b_1 - \cdots - b_k - ac$) where $k = \#\set{b < a}{\omega^{-1}(b) < \omega^{-1}(c)}$.
\end{proposition}

\begin{proof}
This is an immediate consequence of \cref{prop:congruence}: a permutation~$\pi$ is minimal (resp.~maximal) in its $\equiv_\omega$-congruence class if and only if it contains no consecutive exchangeable entries~$ca$ (resp.~$ac$) with~$a < c$.
\end{proof}

\section{Canopy}
\label{sec:canopy}

Recall that the \defn{canopy} of a binary tree~$T$ with~$n$ internal nodes is the sign sequence~${\canopy{T} \in \{{-},{+}\}^{n-1}}$ defined by~$\canopy{T}_i = {-}$ if the $(i+1)$st leaf of~$T$ is a left leaf and~$\canopy{T}_i = {+}$ if the $(i+1)$st leaf of~$T$ is a right leaf.
Equivalently, $\canopy{T}_i = -$ if the node~$i$ of~$T$ is above the node~$i+1$ of~$T$ and~$\canopy{T}_i = +$ otherwise.
This map was already used by J.-L.~Loday in~\cite{LodayRonco, Loday}, but the name ``canopy'' was coined by X.~Viennot~\cite{Viennot}.
% The binary search tree insertion map and the canopy map factorize the recoil map: $\surjectionBrickZono[] \circ \surjectionPermBrick[] = \surjectionPermZono[]$. This combinatorial fact can also be understood on the geometry of the normal fans of the permutahedron, the associahedron and the cube, see \cref{subsec:normalFans}.

Before introducing the canopy, let us define the $\omega$-recoils.
For this, consider the graph~$G(\omega)$ with vertex set~$[n]$ and edge set
\[
\bigset{ij}{i < j \text{ and } j-i \le \#\set{k < i}{\omega^{-1}(k) < \omega^{-1}(j)}}.
\]
Let~$\acyclicOrientations(\omega)$ denote the set of acyclic orientations of~$G(\omega)$.

>>>>

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{notation}[theorem]{Notation}

@@

% pipe dreams
\input{pipeDreamsTikz}
% cross
%\newcommand{\cross}[1][black]{\raisebox{.1cm}{\pipeDreamMonoColor{#1}{black}{c}}}
\newcommand{\cross}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{cross}}}
\newcommand{\crossBiColor}[2]{\raisebox{.1cm}{\pipeDreamBiColor{#1}{#2}{black}{c/l/r}}}
\newcommand{\NScross}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{NScross}}}
\newcommand{\WEcross}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{WEcross}}}
% elbows
%\newcommand{\elbow}[1][black]{\raisebox{.1cm}{\pipeDreamMonoColor{#1}{black}{e}}}
\newcommand{\elbow}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{elbow}}}
\newcommand{\elbowBiColor}[2]{\raisebox{.1cm}{\pipeDreamBiColor{#1}{#2}{black}{e/l/r}}}
%\newcommand{\SEelbow}[1][black]{\raisebox{.1cm}{\pipeDreamBiColor{white}{#1}{black}{e/l/r}}}
\newcommand{\SEelbow}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{SEelbow}}}
%\newcommand{\WNelbow}[1][black]{\raisebox{.1cm}{\pipeDreamBiColor{#1}{white}{black}{e/l/r}}}
\newcommand{\WNelbow}[1][black]{\raisebox{-.15cm}{\includegraphics[scale=.9]{WNelbow}}}
\newcommand{\waving}{\mathsf{wav}}%{\raisebox{-.2cm}{\begin{tikzpicture}[scale=1]\pipeDreamBiColor{.2}{(0,0)}{black}{white}{{t/r/l},{t/l/l},{t/l/r}}\end{tikzpicture}}}
\newcommand{\transpose}{\mathsf{transp}}
\newcommand{\pipeDreams}{\Pi} % set of pipe dreams
\newcommand{\HP}{{\bk\pipeDreams}} % Hopf Algebra of pipe dreams
\newcommand{\reversingPipeDreams}{\Omega} % reversing pipe dreams
\newcommand{\contact}{^\#} % contact graph
\newcommand{\se}{\textsc{se}} % southeast
\newcommand{\nw}{\textsc{nw}} % northwest
\newcommand{\duality}{^\star} % duality pipe dreams - triangulations
\newcommand{\acyclicPipeDreams}{\Sigma} % acyclic pipe dreams
\newcommand{\linearExtensions}{\mathcal{L}} % linear extensions
\newcommand{\noninversions}[2]{\mathsf{ninv}(#1,#2)} % non-inversions
\newcommand{\acyclicOrientations}{\Omega} % set of pipe dreams
\newcommand{\insertion}[2]{\mathsf{pd}(#1,#2)} % insertion of permutation #1 in pipe dreams with permutation #2
\newcommand{\recoils}[2]{\mathsf{rec}(#1,#2)} % recoil scheme of a pipe dream
\newcommand{\canopy}[1]{\mathsf{can}(#1)} % canopy of a pipe dream
\newcommand{\greedyPipeDream}{P^{\overleftarrow{gr}}} % greedy facet
\newcommand{\antiGreedyPipeDream}{P^{\overrightarrow{gr}}} % antigreedy facet

@@

\begin{notation}
For~$\pi \in [e, \omega]$, we denote by~$\insertion{\pi}{\omega}$ the pipe dream of~$\acyclicPipeDreams(\omega)$ such that~${\pi \in \linearExtensions(\insertion{\pi}{\omega})}$.
\end{notation}

@@

We call \defn{pipe dream congruence} the equivalence relation~$\equiv_\omega$ on the weak order interval~$[e,\omega]$ whose equivalence classes are the sets of linear extensions of the pipe dreams of~$\acyclicPipeDreams(\omega)$.
In other words, $\pi \equiv_\omega \pi'$ if and only if~$\insertion{\pi}{\omega} = \insertion{\pi'}{\omega}$.
This indeed defines an equivalence relation by \cref{thm:partitionPipeDreams}.
In this section, we prove the following statement.

@@

\begin{example}
\label{exm:Tamari3}
Following \cref{exm:Tamari1,exm:Tamari2}, observe that the congruence~$\equiv_{\rho_n}$ on the permutations of~$\{0, 1, \dots, n, n+1\}$ below~$\rho_n$ corresponds to the sylvester congruence on~$\fS_n$.
The classes of this congruence are the sets of linear extensions of the binary trees (considered as posets, labeled in inorder, and oriented towards their roots).
It can also be defined by the classical rewriting rule~$UjVikW \equiv UjVkiW$ where~$i < j < k$ are elements of~$[n]$ while~$U, V, W$ are (possibly empty) words on~$[n]$.
\end{example}

@@

In this section, we discuss four further topics on the pipe dream congruence.
In \cref{subsec:sweepingAlgorithm,subsec:insertionAlgorithm}, we describe two algorithms to construct the pipe dream~$\insertion{\pi}{\omega}$ of~$\acyclicPipeDreams(\omega)$ of which a given permutation~$\pi$ is a linear extension.
In \cref{subsec:rewritingRule}, we describe the pipe dream congruence~$\equiv_\omega$ as the transitive closure of a rewriting rule on permutations of~$[e, \omega]$.
Finally, in \cref{subsec:canopy}, we describe the natural coarsening of the pipe dream congruence~$\equiv_\omega$ by the recoil congruence~$\cong_\omega$.

@@

Our first algorithm to construct~$\insertion{\pi}{\omega}$ is a \defn{sweeping algorithm}, inspired by the algorithm to compute greedy pipe dreams~\cite{PilaudPocchiola, PilaudStump-ELlabelings}.
Extensions of this algorithm will be discussed in \cref{..}. \vincent{todo. Was this algorithm explicitly somewhere before? We should cite Jahn-Stump later in the paper.}

@@

Our second algorithm to construct~$\insertion{\pi}{\omega}$ is an \defn{insertion algorithm} inspired from~\cite{Pilaud-BrickAlgebra} and similar to the insertion in binary search trees.

@@

\begin{lemma}
\label{lem:rectangleInsertionAlgorithm}
For any~$t, r, c \in [n]$, the free southeast elbows in the rectangle~$[r] \times [c]$ just before step~$t$ of the insertion algorithm form a staircase of length
\[
\# \set{s < t}{\pi(s) \le r \text{ and } \omega^{-1}(\pi(s)) \le c} - \# \set{s < t}{\pi(s) > r \text{ and } \omega^{-1}(\pi(s)) > c}.
\]
\end{lemma}

\begin{proof}
Denote by~$R$ the rectangle~$[r] \times [c]$.
The proof works by induction on~$t$.
Before step~$1$, there is no southeast elbow in~$R$.
Assume now that just before step~$t$, the free southeast elbows in~$R$ form a staircase~$e_1, \dots, e_k$ with~$k$ given by the formula of the statement.
At step~$t$, we insert a new pipe~$\pi(t)$ which enters at~$\pi(t)$ and exits at~$\omega^{-1}(\pi(t))$.
Let~${0 \le i < j \le k+1}$ be such that~$e_{i+1}, \dots, e_{j-1}$ are the northwest elbows of pipe~$\pi(t)$ in~$R$, and let~$e'_1, \dots, e'_\ell$ be the southeast elbows of pipe~$\pi(t)$ in~$R$.
Hence, the free southeast elbows in~$R$ after step~$t$ form the sequence~$e_1, \dots, e_i, e'_1, \dots, e'_\ell, e_j, \dots, e_k$.
We thus just need to see that $e_i$ is southwest of~$e'_1$ and~$e'_\ell$ is southwest of~$e_j$, and that
\[
\ell = \begin{cases} j-i  & \text{if } \pi(s) \le r \text{ and } \omega^{-1}(\pi(s)) \le c, \\  j-i-2  & \text{if } \pi(s) > r \text{ and } \omega^{-1}(\pi(s)) > c, \\  j-i-1 & \text{otherwise.} \end{cases}
\]
Since the northwest and southeast elbows alternate along pipe~$\pi(t)$, this follows from the fact that the pipe~$\pi(t)$ enters~$R$ with an horizontal step if~$\pi(t) \le r$ and with a vertical step if~$\pi(s) > r$, and exits~$R$ with a vertical step if~$\omega^{-1}(\pi(t)) \le c$ and with an horizontal step if~$\omega^{-1}(\pi(s)) > c$.
%setting~$i \eqdef \pi(t)$ and~$j \eqdef \pi(s)$, we have~${i < j}$ and~$\pi^{-1}(i) > \pi^{-1}(j)$ while~$\omega^{-1}(i) < \omega^{-1}(j)$, which contradicts our assumption that~$\pi < \omega$. \qedhere
\end{proof}

@@

\begin{proof}
Observe first that if there if~$r < s$ such that~$\pi(r) > \pi(s)$ and~$\omega^{-1}(\pi(r)) > \omega^{-1}(\pi(s))$, then setting~$i \eqdef \pi(s)$ and~$j \eqdef \pi(r)$, we have~${i < j}$ and~$\pi^{-1}(i) > \pi^{-1}(j)$ while~$\omega^{-1}(i) < \omega^{-1}(j)$, which contradicts our assumption that~$\pi < \omega$.
This implies that
\begin{align*}
\# \set{s < t}{\pi(s) < \pi(t) \text{ and } \omega^{-1}(\pi(s)) < \omega^{-1}(\pi(t))} & = \noninversions{\omega}{\pi(t)} \\
\text{and}\qquad \# \set{s < t}{\pi(s) > \pi(t) \text{ and } \omega^{-1}(\pi(s)) > \omega^{-1}(\pi(t))} & = 0.
\end{align*}
Hence, the result follows from~\cref{lem:rectangleInsertionAlgorithm} applied to the parameters~$t, \pi(t), \omega^{-1}(\pi(t))$.
\end{proof}

@@

@@

@@

Recall from \cref{exm:Tamari3} that the sylvester congruence can be defined as the transitive closure of the classical rewriting rule~$UjVikW \equiv UjVkiW$ where~$i < j < k$ are elements of~$[n]$ while~$U, V, W$ are (possibly empty) words on~$[n]$ (as usual, we write the permutations of~$\fS_n$ as words in one-line notation).
We now describe a similar rewriting rule for the pipe dream congruence~$\equiv_\omega$.

\begin{proposition}
\label{prop:rewritingRule}
On the interval~$[e,\omega]$ of the weak order, the pipe dream congruence~$\equiv_\omega$ coincides with the transitive closure of the rewriting rule~$U ij V \equiv_\omega U ji V$ where~$1 \le i < j \le n$ are elements of~$[n]$ while~$U, V$ are (possibly empty) words on~$[n]$ such that
\[
\# \set{k \in U}{k > i} \ge \# \set{k \in U}{\omega^{-1}(k) < \omega^{-1}(j)}.
\]
\end{proposition}

\begin{proof}
As they are linear extensions of posets, the congruence classes of $\equiv_\omega$ are connected by simple transpositions.
We thus just need to show that any two permutations~$\pi \eqdef UijV$ and~$\pi' \eqdef UjiV$ of~$[e, \omega]$ which differ by the inversion of two consecutive values are equivalent for~$\equiv_\omega$ if and only if~$\# \set{k \in U}{k > i} \ge \# \set{k \in U}{\omega^{-1}(k) < \omega^{-1}(j)}$.
Moreover, by \cref{prop:insertionAlgorithm}, $\pi \equiv_\omega \pi'$ if and only if they are sent to the same pipe dream by the insertion algorithm.
Let~$t \eqdef \pi^{-1}(i) = \pi'^{-1}(j)$.
Before step~$t$ of the insertion algorithm, we insert the word~$U$ both for~$\pi$ and for~$\pi'$.
The insertion of~$i$ and~$j$ then commute if and only if there is no currently free elbow in the rectangle~$[i] \times [\omega^{-1}(j)]$.
By \cref{lem:rectangleInsertionAlgorithm} applied to the parameters~$t, i, \omega^{-1}(j)$, this is equivalent to
\[
\# \set{s < t}{\pi(s) \le i \text{ and } \omega^{-1}(\pi(s)) \le \omega^{-1}(j)} \le \# \set{s < t}{\pi(s) > i \text{ and } \omega^{-1}(\pi(s)) > \omega^{-1}(j)}.
\]
or written differently,
\[
\# \set{k \in U}{k < i \text{ and } \omega^{-1}(k) < \omega^{-1}(j)} \le \# \set{k \in U}{k > i \text{ and } \omega^{-1}(k) > \omega^{-1}(j)}.
\]
(note that  to replace large by strict inequalities in the first set, we used that~$k \ne i$ and ${\omega^{-1}(k) \ne \omega^{-1}(j)}$ since~$i,j \notin U$ while~$k \in U$).
Finally, observe that adding~$\# \set{k \in U}{k > i \text{ and } \omega^{-1}(k) < \omega^{-1}(j)}$ from both side, we obtain the equivalent condition
\[
\# \set{k \in U}{\omega^{-1}(k) < \omega^{-1}(j)} \le \# \set{k \in U}{k > i}.
\qedhere
\]
\end{proof}

\begin{remark}
\label{rem:rewritingRule}
Observe from the proof that the condition of \cref{prop:rewritingRule} is equivalent to
\[
\# \set{k \in U}{k < i \text{ and } \omega^{-1}(k) < \omega^{-1}(j)} \le \# \set{k \in U}{k > i \text{ and } \omega^{-1}(k) > \omega^{-1}(j)}.
\]
Observe moreover that since~$\pi \le \omega$ and~$i < j$, we have
\begin{align*}
\set{k \in U}{k < i \text{ and } \omega^{-1}(k) < \omega^{-1}(j)} & = \set{k \in [i]}{\omega^{-1}(k) < \omega^{-1}(j)} \qquad \text{and}\\
\set{k \in U}{k > i \text{ and } \omega^{-1}(k) > \omega^{-1}(j)} & = \set{k \in [n]}{i < k < j \text{ and } \omega^{-1}(i) > \omega^{-1}(k) > \omega^{-1}(j)}.
\end{align*}
\end{remark}

We close this section by two immediate consequences of \cref{prop:rewritingRule,rem:rewritingRule}.

\begin{corollary}
\label{coro:patternAvoiding}
A permutation is minimal (resp.~maximal) in its pipe dream congruence class if and only if it avoids the patters~$k_1 - \cdots - k_p - ji$ (resp.~$k_1 - \cdots - k_p - ij$) where~$k_q > i$ and~$\omega^{-1}(k_q) > \omega^{-1}(j)$ for~$q \in [p]$ and~$p = \# \set{k \in [i]}{\omega^{-1}(k) < \omega^{-1}(j)}$.
\end{corollary}

\begin{proof}
This is an immediate consequence of \cref{prop:rewritingRule}: a permutation~$\pi$ is minimal (resp.~maximal) in its $\equiv_\omega$-congruence class if and only if it contains no consecutive exchangeable entries~$ji$ (resp.~$ij$) with~$i < j$.
\end{proof}

\begin{corollary}
\label{coro:canopy}
If~$1 \le i < j \le n$ and~$j-i \le \#\set{k \in [i]}{\omega^{-1}(k) < \omega^{-1}(j)}$, then the pipes~$i$ and~$j$ are comparable for $\contactLess{P}$ in any acyclic pipe dream~$P \in \acyclicPipeDreams(\omega)$.
\end{corollary}

\begin{proof}
If pipes~$i$ and~$j$ were incomparable in~$\contactLess{P}$, there would be two permutations~$UijV \equiv_\omega UjiV$.
However, $j-i \le \#\set{k \in [i]}{\omega^{-1}(k) < \omega^{-1}(j)}$ implies that the condition of \cref{prop:rewritingRule} cannot hold, by \cref{rem:rewritingRule}
\end{proof}

\subsection{Recoils and canopy}
\label{subsec:canopy}

To conclude, we generalize the notion of recoils of permutations and of canopy of a binary trees to show a natural commutative diagram of lattice homomorphisms.
We start with the easy generalization of recoils of a permutation.

\begin{definition}
\label{def:recoils}
Consider the graph~$G(\omega)$ with vertex set~$[n]$ and edge set
\[
\set{ik}{i < j \text{ and } j-i \le \#\set{k \in [i]}{\omega^{-1}(k) < \omega^{-1}(j)}}.
\]
Let~$\acyclicOrientations(\omega)$ denote the set of acyclic orientations of~$G(\omega)$.
The \defn{$\omega$-recoil congruence} is the equivalence relation on the weak order interval~$[e, \omega]$ whose equivalence classes are the sets of linear extensions of the acyclic orientations of~$G(\omega)$.
The \defn{$\omega$-recoils} of a permutation~$\pi \in [e,\omega]$ is the acyclic orientation~${\recoils{\pi}{\omega} \in \acyclicOrientations(\omega)}$ such that~$\pi$ is a linear extension of~$\recoils{\pi}{\omega}$.
\end{definition}
